### 2021-05-23 content/80.formal-definition.tex
where have all the DOIs gone..

\begin{eqnarray*}
    \mathcal{L}_{rocrate}   & \equiv & \big\{ Property(p), Class(c), 
                            Value(x), \mathbb{R}, \mathbb{S} \big\} \\
    \mathbb{D}              & \equiv & \mathbb{IRI} \\
    \mathbb{IRI}            & \equiv & { \text{IRIs as defined in RFC3987} } \\
    \mathbb{R}              & \equiv & { \text{real or integer numbers} } \\
    \mathbb{S}              & \equiv & { \text{literal strings} }
\end{eqnarray*}

\begin{markdown}
The domain of discourse $\mathbb{D}$ is the set of $\mathbb{IRI}$ identifiers [@doi:10.17487/rfc3987] (notation `<http://example.com/>`)[^3], with additional descriptions using numbers $\mathbb{R}$ (notation $13.37$) and literal strings $\mathbb{S}$ (notation $\text{“Hello”}$).

@@

[^3]:
    For simplicity, blank nodes are not included in this formalisation, as RO-Crate
    recommends the use of IRI identifiers: \url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html#describing-entities-in-json-ld} 
\end{markdown}

\begin{markdown}
### Minimal RO-Crate

Below we use $\mathcal{L}_{rocrate}$ to define a minimal[^4] RO-Crate:

[^4]: The full list of types, relations and attribute properties from the RO-Crate specification are not included. Examples shown include $datePublished$, $CreativeWork$ and $name$.
\end{markdown}

@@

\begin{markdown}
An $ROCrate(R)$ is defined as a self-described _Root Data Entity_, which describes and contains parts (_data entities_), which are further described in _contextual entities_.  These terms align with their use in the [RO-Crate 1.1 terminology](https://www.researchobject.org/ro-crate/1.1/terminology).

The $Root(r)$ is a type of $Dataset(r)$, and must have the metadata to literal attributes to provide a $name$, $description$ and $datePublished$, as well as a contextual entity identifying its license. These predicates correspond to the RO-Crate 1.1 [requirements for the root data entity](https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#direct-properties-of-the-root-data-entity).

The concept of an $Entity(e)$ is introduced as being either a $DataEntity(e)$, a $ContextualEntity(e)$, or [both](https://www.researchobject.org/ro-crate/1.1/contextual-entities.html#contextual-vs-data-entities); and must be typed with at least one $Class(e)$.

For simplicity in this formalization (and to assist production rules below) $R$ is a constant representing a single RO-Crate, typically written to independent RO-Crate Metadata files. $R$ is used by $Mentions(R, e)$ to indicate that $e$ is an Entity described by the RO-Crate and therefore its metadata (a set of $Relation$ and $Attribute$ predicates) form part of the RO-Crate serialization. $Relation(s, p, o)$ and $Attribute(s, p, x)$ are defined as a _subject-predicate-object_ triple pattern from an $Entity(s)$ using a $Property(p)$ to either another $Entity(o)$ or a $Value(x)$ value.
\end{markdown}

\begin{markdown}
### Example of formalized RO-Crate

The below is an example RO-Crate represented using the above formalization, assuming a base URI of `<http://example.com/ro/123/>`:
\end{markdown}

@@

\begin{markdown}
In reality many additional attributes from schema.org types like <http://schema.org/Dataset> and <http://schema.org/CreativeWork> would be used to further describe the RO-Crate and its entities, but as these are optional they do not form part of this formalization.

### Mapping to RDF with schema.org

A formalized RO-Crate in 
$\mathcal{L}_{rocrate}$
can be mapped to different serializations.

Assume a simplified[^4] language 
$\mathcal{L}_{RDF}$ 
based on the [RDF abstract syntax](http://www.w3.org/TR/2014/REC-rdf11-concepts-20140225/):

[^4]: This simplification does not cover the extensive list of literal datatypes built-in to RDF 1.1, only strings and decimal real numbers. Likewise, language of literals are not included.

\end{markdown}

@@

\begin{markdown}
Note that in the JSON-LD serialization of RO-Crate, the expression of $Class$ and $Property$ is typically indirect: The JSON-LD `@context` maps to schema.org IRIs, which, when resolved as Linked Data, embed their formal definition as RDFa.

### RO-Crate 1.1 Metadata File Descriptor

An important RO-Crate principle is that of being **self-describing**. Therefore the serialization of the RO-Crate into a file should also describe itself in a [Metadata File Descriptor](https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#ro-crate-metadata-file-descriptor), indicating it is about (describing) the RO-Crate root data entity, and that it conformsTo a particular version of the RO-Crate specification:
\end{markdown}

@@

\begin{markdown}
Note that although the metadata file necessarily is an _information resource_ written to disk or served over the network (e.g. as JSON-LD), it is not considered to be a contained _part_ of the RO-Crate in the form of a _data entity_, rather it is described only as a _contextual entity_.

While in the conceptual model the _RO-Crate Metadata File_ can be seen as the top-level node that describes the _RO-Crate Root_, in the formal model (and the JSON-LD format) the metadata file descriptor is an additional contextual entity and not affecting the depth-limit of the RO-Crate.
\end{markdown}

\begin{markdown}
### Forward-chained Production Rules for JSON-LD

Combining the above predicates and schema.org mapping with rudimentary JSON templates, these forward-chaining production rules can output JSON-LD according to the RO-Crate 1.1 specification[^2]:

[^2]:
    **Limitations:** Contextual entities not related from the RO-Crate (e.g. using inverse relations to a data entity) would not be covered by the single direction $Mentions(R, s)$ production rule; see [issue 122](https://github.com/ResearchObject/ro-crate/issues/122). The $datePublished(e, date)$ rule do not include syntax checks for the ISO 8601 datetime format. Compared with RO-Crate examples, this generated JSON-LD does not use a $@context$ as the IRIs are produced unshortened, a post-step could do JSON-LD Flattening with a versioned RO-Crate context. The `@type` expansion is included for clarity, even though this is also implied by the $type(e, t)$ expansion to $Relation(e, \texttt{xsd:type})$.
\end{markdown}

@@

\begin{markdown}
This exposes the first order logic domain of discourse of IRIs, with rational numbers and strings as their corresponding JSON-LD representation. These production rules first grow the graph of $R$ by adding a transitive rule – anything described in $R$ which is related to $o$, means that $o$ is also mentioned by the $ROCrate(R)$. For simplicity this rule is one-way; in practice the graph can also contain free-standing contextual entities that have outgoing relations to data- and contextual entities.
\end{markdown}

>>>>

\begin{eqnarray*}
    \mathcal{L}_{rocrate}   & = & \big\{ Property(p), Class(c), 
                            Value(x), \mathbb{R}, \mathbb{S} \big\} \\
    \mathbb{D}              & = & \mathbb{IRI} \\
    \mathbb{IRI}            & \equiv & { \text{IRIs as defined in RFC3987} } \\
    \mathbb{R}              & \equiv & { \text{real or integer numbers} } \\
    \mathbb{S}              & \equiv & { \text{literal strings} }
\end{eqnarray*}

The domain of discourse $\mathbb{D}$ is the set of $\mathbb{IRI}$ identifiers \cite{doi:10.17487/rfc3987} (notation $\texttt{<http://example.com/>}$)\footnote{
    For simplicity, blank nodes are not included in this formalisation, as RO-Crate
    recommends the use of IRI identifiers: \url{https://www.researchobject.org/ro-crate/1.1/appendix/jsonld.html#describing-entities-in-json-ld} 
}, with additional descriptions using numbers $\mathbb{R}$ (notation $13.37$) and literal strings $\mathbb{S}$ (notation $\text{“Hello”}$).

@@

\subsection{Minimal RO-Crate}

Below we use $\mathcal{L}_{rocrate}$ to define a minimal\footnote{
    The full list of types, relations and attribute properties from the RO-Crate specification are not included. Examples shown include $datePublished$, $CreativeWork$ and $name$.
} RO-Crate:

@@

An $ROCrate(R)$ is defined as a self-described \emph{Root Data Entity}, which describes and contains parts (\emph{data entities}), which are further described in \emph{contextual entities}.  These terms align with their use in the \href{https://www.researchobject.org/ro-crate/1.1/terminology}{RO-Crate 1.1 terminology}.

The $Root(r)$ is a type of $Dataset(r)$, and must have the metadata to literal attributes to provide a $name$, $description$ and $datePublished$, as well as a contextual entity identifying its license. These predicates correspond to the RO-Crate 1.1 \href{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#direct-properties-of-the-root-data-entity}{requirements for the root data entity}.

The concept of an $Entity(e)$ is introduced as being either a $DataEntity(e)$, a $ContextualEntity(e)$, or \href{https://www.researchobject.org/ro-crate/1.1/contextual-entities.html#contextual-vs-data-entities}{both}; and must be typed with at least one $Class(e)$.

For simplicity in this formalization (and to assist production rules below) $R$ is a constant representing a single RO-Crate, typically written to independent RO-Crate Metadata files. $R$ is used by $Mentions(R, e)$ to indicate that $e$ is an Entity described by the RO-Crate and therefore its metadata (a set of $Relation$ and $Attribute$ predicates) form part of the RO-Crate serialization. $Relation(s, p, o)$ and $Attribute(s, p, x)$ are defined as a \emph{subject—predicate—object} triple pattern from an $Entity(s)$ using a $Property(p)$ to either another $Entity(o)$ or a $Value(x)$ value.

\subsection{Example of formalized RO-Crate}

The below is an example RO-Crate represented using the above formalisation, assuming a base URI of \texttt{<http://example.com/ro/123/>}:

@@

In reality many additional attributes from schema.org types like \url{http://schema.org/Dataset} and \url{http://schema.org/CreativeWork} would be used to further describe the RO-Crate and its entities, but as these are optional they do not form part of this formalisation.

\subsection{Mapping to RDF with schema.org}

A formalized RO-Crate in $\mathcal{L}_{rocrate}$ can be mapped to different serializations.
Assume a simplified\footnote{
 This simplification does not cover the extensive list of literal datatypes built-in to RDF 1.1, only strings and decimal real numbers. Likewise, language of literals are not included.
} language $\mathcal{L}_{RDF}$ 
based on the \href{http://www.w3.org/TR/2014/REC-rdf11-concepts-20140225/}{RDF abstract syntax}:

@@

Note that in the JSON-LD serialization of RO-Crate, the expression of $Class$ and $Property$ is typically indirect: The JSON-LD \texttt{@context} maps to schema.org IRIs, which, when resolved as Linked Data, embed their formal definition as RDFa.

\subsection{RO-Crate 1.1 Metadata File Descriptor}

An important RO-Crate principle is that of being \textbf{self-describing} Therefore the serialisation of the RO-Crate into a file should also describe itself in a \href{https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#ro-crate-metadata-file-descriptor}{Metadata File Descriptor}, indicating it is $about$ (describing) the RO-Crate root data entity, and that it $conformsTo$ a particular version of the RO-Crate specification:

@@

Note that although the metadata file necessarily is an \emph{information resource} written to disk or served over the network (e.g. as JSON-LD), it is not considered to be a contained \emph{part} of the RO-Crate in the form of a \emph{data entity}, rather it is described only as a \emph{contextual entity}.

While in the conceptual model the \emph{RO-Crate Metadata File} can be seen as the top-level node that describes the \emph{RO-Crate Root}, in the formal model (and the JSON-LD format) the metadata file descriptor is an additional contextual entity and not affecting the depth-limit of the RO-Crate.

\subsection{Forward-chained Production Rules for JSON-LD}

Combining the above predicates and schema.org mapping with rudimentary JSON templates, these forward-chaining production rules can output JSON-LD according to the RO-Crate 1.1 specification\footnote{
    \textbf{Limitations:} Contextual entities not related from the RO-Crate (e.g. using inverse relations to a data entity) would not be covered by the single direction $Mentions(R, s)$ production rule; see \href{https://github.com/ResearchObject/ro-crate/issues/122}{issue 122}. The $datePublished(e, date)$ rule do not include syntax checks for the ISO 8601 datetime format. Compared with RO-Crate examples, this generated JSON-LD does not use a $@context$ as the IRIs are produced unshortened, a post-step could do JSON-LD Flattening with a versioned RO-Crate context. The \texttt{@type} expansion is included for clarity, even though this is also implied by the $type(e, t)$ expansion to $Relation(e, \texttt{xsd:type})$.
}:

@@

This exposes the first order logic domain of discourse of IRIs, with rational numbers and strings as their corresponding JSON-LD representation. These production rules first grow the graph of $R$ by adding a transitive rule – anything described in $R$ which is related to $o$, means that $o$ is also mentioned by the $ROCrate(R)$. For simplicity this rule is one-way; in practice the graph can also contain free-standing contextual entities that have outgoing relations to data- and contextual entities.
